skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-07-13 paper/paper.Rmd
vpn edits to skater paper

motivation: |
  SNP-based kinship analysis with genome-wide relationship estimation and IBD segment analysis methods produce results that often require further downstream manipulation and analysis. A software package with a consistent design and implementation that offers this downstream analysis functionality is needed.

results: |
  Here we present the skater package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation as an **R** package. The skater package contains a suite of well-documented tools for importing, parsing, and analyzing pedigree data, performing relationship degree inference, benchmarking relationship degree classification, and summarizing IBD segment data.

@@

Inferring familial relationships between individuals using genetic data is a common problem in population genetics, medical genetics, and forensics. There are multiple approaches to estimating the relatedness between samples, including genome-wide relatedness measures such as those implemented in Plink [@purcell2007] or KING [@manichaikul2010], and methods that rely on identity by descent (IBD) segment detection such as GERMLINE [@gusev2009], hap-IBD [@zhou2020], and IBIS [@seidman2020]. Recent efforts have focused on benchmarking these methods [@ramstetter2017; @vries2021] aided by tools for simulating pedigrees and genome-wide SNP data [@caballero2019]. Analyzing results from genome-wide SNP-based kinship analysis or comparing analyses to simulated data for benchmarking have to this point required writing one-off analysis functions or utility scripts that are rarely shared and/or poorly documented. There is a need in the field for a well-documented software package with a consistent design and API that contains functions to assist with downstream manipulation, benchmarking, and analysis of SNP-based kinship assessment methods. Here we present the skater R package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation.

@@

The skater package provides a tidyverse-friendly collection of analysis and utility functions for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation as an **R** package. Functions in the package include tools for importing, parsing, and analyzing pedigree data, performing relationship degree inference, benchmarking relationship degree classification, and summarizing IBD segment data.

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the functions `read_fam()` to read in a PLINK-formatted .fam file, and another function `fam2ped()` to convert these into an internal pedigree object as a nested tibble with one row per family. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to truth. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette (see Supplementary Material), and is a superior measure to classification accuracy when benchmarking relationship degree estimation. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second degree) than misclassifications at more distant relationships (e.g., misclassifying a fourth-degree relative pair as fifth-degree).

@@

The skater R package provides a thoroughly-documented software package for common data import, manipulation, and analysis tasks typically encountered when working with common SNP-based kinship analysis tools. All package functions are internally documented with examples, and the package contains a vignette demonstrating usage, inputs, outputs, and interpretation of all key functions (see Supplementary Material). The package contains internal tests which are automatically run with continuous integration via GitHub Actions whenever the package code is updated. The skater package is permissively licensed (MIT) and is easily extensible to accommodate outputs from new genome-wide relatedness and IBD segment methods as they become available.

>>>>

motivation: |
  SNP-based kinship analysis with genome-wide relationship estimation and IBD segment analysis methods produces results that often require further downstream processing and manipulation. A dedicated software package that consistently and intuitively implements this analysis functionality is needed.

results: |
  Here we present the skater R package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation with **R**. The skater package contains a suite of well-documented tools for importing, parsing, and analyzing pedigree data, performing relationship degree inference, benchmarking relationship degree classification, and summarizing IBD segment data.

@@

Inferring familial relationships between individuals using genetic data is a common problem in population genetics, medical genetics, and forensics. There are multiple approaches to estimating relatedness between samples, including genome-wide measures such as those implemented in Plink [@purcell2007] or KING [@manichaikul2010], and methods that rely on identity by descent (IBD) segment detection such as GERMLINE [@gusev2009], hap-IBD [@zhou2020], and IBIS [@seidman2020]. Recent efforts focusing on benchmarking these methods [@ramstetter2017; @vries2021] have been aided by tools for simulating pedigrees and genome-wide SNP data [@caballero2019]. Analyzing results from genome-wide SNP-based kinship analysis or comparing analyses to simulated data for benchmarking have to this point required writing one-off analysis functions or utility scripts that are seldom distributed with robust documentation, test suites, or narrative examples of usage. There is a need in the field for a well-documented software package with a consistent design and API that contains functions to assist with downstream manipulation, benchmarking, and analysis of SNP-based kinship assessment methods. Here we present the skater package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation with **R**.

@@

The skater package provides an intuitive collection of analysis and utility functions for SNP-based kinship analysis. Functions in the package include tools for importing, parsing, and analyzing pedigree data, performing relationship degree inference, benchmarking relationship degree classification, and summarizing IBD segment data. The package is designed to adhere to "tidy" data analysis principles, and builds upon the tools released under the tidyverse R ecosystem [@Wickham2019].

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the function `read_fam()` to read in a PLINK-formatted .fam file, and another function `fam2ped()` to convert the content into a pedigree object as a nested tibble with one row per family. All pedigree processing from skater internally leverages a data structure from the kinship2 package [@kinship2]. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to truth. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette (see Supplementary Material), and may be a preferrable measure of classification accuracy when benchmarking relationship degree estimation. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second degree) than misclassifications at more distant relationships (e.g., classifying a fourth-degree relative pair as fifth-degree).

@@

The skater R package provides a robust software package for data import, manipulation, and analysis tasks typically encountered when working with SNP-based kinship analysis tools. All package functions are internally documented with examples, and the package contains a vignette demonstrating usage, inputs, outputs, and interpretation of all key functions (see Supplementary Material). The package contains internal tests that are automatically run with continuous integration via GitHub Actions whenever the package code is updated. The skater package is permissively licensed (MIT) and is easily extensible to accommodate outputs from new genome-wide relatedness and IBD segment methods as they become available.

### 2021-07-13 paper/paper.tex
vpn edits to skater paper

\access{Advance Access Publication Date:   }
\appnotes{Applications Note}

@@

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1}, V.P.
Nagraj\,\textsuperscript{1}, Matthew
Scholz\,\textsuperscript{1}, Shakeel Jessa\,\textsuperscript{1}, Carlos
Acevedo\,\textsuperscript{1}, Jianye Ge\,\textsuperscript{2}, August
Woener\,\textsuperscript{2}, Bruce Budowle\,\textsuperscript{2}
}

@@

\abstract{
\textbf{Motivation:} SNP-based kinship analysis with genome-wide
relationship estimation and IBD segment analysis methods produce results
that often require further downstream manipulation and analysis. A
software package with a consistent design and implementation that offers
this downstream analysis functionality is needed.\\
\textbf{Results:} Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation as an \textbf{R} package. The skater package
contains a suite of well-documented tools for importing, parsing, and
analyzing pedigree data, performing relationship degree inference,
benchmarking relationship degree classification, and summarizing IBD
segment data.\\
\textbf{Availability:} The skater package is implemented as an R package
and is released under the MIT license at
https://github.com/signaturescience/skater. Documentation is available
at https://signaturescience.github.io/skater.\\
\textbf{Contact:}sturner@signaturescience.com\\
\textbf{Supplementary information:} Supplementary data are available at
Bioinformatics Online.}

@@

Inferring familial relationships between individuals using genetic data
is a common problem in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating the relatedness
between samples, including genome-wide relatedness measures such as
those implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts have
focused on benchmarking these methods \citep{ramstetter2017, vries2021}
aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are rarely shared and/or poorly
documented. There is a need in the field for a well-documented software
package with a consistent design and API that contains functions to
assist with downstream manipulation, benchmarking, and analysis of
SNP-based kinship assessment methods. Here we present the skater R
package for \textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis,
\textbf{t}esting, and \textbf{e}valuation.

@@

The skater package provides a tidyverse-friendly collection of analysis
and utility functions for \textbf{S}NP-based \textbf{k}inship
\textbf{a}nalysis, \textbf{t}esting, and \textbf{e}valuation as an
\textbf{R} package. Functions in the package include tools for
importing, parsing, and analyzing pedigree data, performing relationship
degree inference, benchmarking relationship degree classification, and
summarizing IBD segment data.

@@

The skater package has several functions for importing, parsing, and
analyzing pedigree data. Pedigrees define familial relationships in a
hierarchical structure. Many genomics tools for working with pedigrees
start with a .fam file, which is a tabular format with one row per
individual and columns for unique IDs of the mother, father, and the
family unit. The skater package contains the functions
\texttt{read\_fam()} to read in a PLINK-formatted .fam file, and another
function \texttt{fam2ped()} to convert these into an internal pedigree
object as a nested tibble with one row per family. Further functions
such as \texttt{plot\_pedigree()} produce a multi-page PDF drawing a
diagram of the pedigree for each family, while \texttt{ped2kinpair()}
produces a pairwise list of relationships between all individuals in the
data with the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to truth. When aggregated over many
relationships and inferences, this can help benchmark performance of a
particular kinship analysis method. The skater package adapts a
\texttt{confusion\_matrix()} function from \citet{clark2021} to provide
standard contingency table metrics (e.g.~sensitivity, specificity, PPV,
precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric.
The R-RMSE metric is defined more thoroughly in the skater package
vignette (see Supplementary Material), and is a superior measure to
classification accuracy when benchmarking relationship degree
estimation. Taking the reciprocal of the target and predicted degree in
a typical RMSE calculation results in larger penalties for more
egregious misclassifications (e.g., classifying a first-degree relative
pair as second degree) than misclassifications at more distant
relationships (e.g., misclassifying a fourth-degree relative pair as
fifth-degree).

>>>>

\access{Advance Access Publication Date:   }
\appnotes{Manuscript Category}

@@

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1},
V.P. Nagraj\,\textsuperscript{1},
Matthew Scholz\,\textsuperscript{1},
Shakeel Jessa\,\textsuperscript{1},
Carlos Acevedo\,\textsuperscript{1},
Jianye Ge\,\textsuperscript{2},
August Woener\,\textsuperscript{2},
Bruce Budowle\,\textsuperscript{2},
}

@@

\abstract{
\textbf{Motivation:} SNP-based kinship analysis with genome-wide relationship estimation and
IBD segment analysis methods produces results that often require further
downstream processing and manipulation. A dedicated software package
that consistently and intuitively implements this analysis functionality
is needed.\\
\textbf{Results:} Here we present the skater R package for \textbf{S}NP-based
\textbf{k}inship \textbf{a}nalysis, \textbf{t}esting, and
\textbf{e}valuation with \textbf{R}. The skater package contains a suite
of well-documented tools for importing, parsing, and analyzing pedigree
data, performing relationship degree inference, benchmarking
relationship degree classification, and summarizing IBD segment data.\\
\textbf{Availability:} The skater package is implemented as an R package and is released under
the MIT license at https://github.com/signaturescience/skater.
Documentation is available at https://signaturescience.github.io/skater.\\
\textbf{Contact:}sturner@signaturescience.com\\
\textbf{Supplementary information:} Supplementary data are available at Bioinformatics Online.}

@@

Inferring familial relationships between individuals using genetic data
is a common problem in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating relatedness
between samples, including genome-wide measures such as those
implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts focusing
on benchmarking these methods \citep{ramstetter2017, vries2021} have
been aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are seldom distributed with robust
documentation, test suites, or narrative examples of usage. There is a
need in the field for a well-documented software package with a
consistent design and API that contains functions to assist with
downstream manipulation, benchmarking, and analysis of SNP-based kinship
assessment methods. Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation with \textbf{R}.

@@

The skater package provides an intuitive collection of analysis and
utility functions for SNP-based kinship analysis. Functions in the
package include tools for importing, parsing, and analyzing pedigree
data, performing relationship degree inference, benchmarking
relationship degree classification, and summarizing IBD segment data.
The package is designed to adhere to ``tidy'' data analysis principles,
and builds upon the tools released under the tidyverse R ecosystem
\citep{Wickham2019}.

@@

The skater package has several functions for importing, parsing, and
analyzing pedigree data. Pedigrees define familial relationships in a
hierarchical structure. Many genomics tools for working with pedigrees
start with a .fam file, which is a tabular format with one row per
individual and columns for unique IDs of the mother, father, and the
family unit. The skater package contains the function
\texttt{read\_fam()} to read in a PLINK-formatted .fam file, and another
function \texttt{fam2ped()} to convert the content into a pedigree
object as a nested tibble with one row per family. All pedigree
processing from skater internally leverages a data structure from the
kinship2 package \citep{kinship2}. Further functions such as
\texttt{plot\_pedigree()} produce a multi-page PDF drawing a diagram of
the pedigree for each family, while \texttt{ped2kinpair()} produces a
pairwise list of relationships between all individuals in the data with
the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to truth. When aggregated over many
relationships and inferences, this can help benchmark performance of a
particular kinship analysis method. The skater package adapts a
\texttt{confusion\_matrix()} function from \citet{clark2021} to provide
standard contingency table metrics (e.g.~sensitivity, specificity, PPV,
precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric.
The R-RMSE metric is defined more thoroughly in the skater package
vignette (see Supplementary Material), and may be a preferrable measure
of classification accuracy when benchmarking relationship degree
estimation. Taking the reciprocal of the target and predicted degree in
a typical RMSE calculation results in larger penalties for more
egregious misclassifications (e.g., classifying a first-degree relative
pair as second degree) than misclassifications at more distant
relationships (e.g., classifying a fourth-degree relative pair as
fifth-degree).
